---
title: "3: Propensity scores"
subtitle: "r4arm"
author: Andrea Mazzella [github](https://github.com/andreamazzella)
output: html_notebook
---

# What you will learn

* calculate a propensity score
* estimate a treatment effect
* propensity score matching

-------------------------------------------------------------------------------

# Data
`RFA_pscore.dta`, via LSHTM Moodle (not publicly available).

Key points:
* Simulated dataset from a hypothetical cohort study
* Patients with lung metastases
* Treatment `rfa`, either radiofrequency ablation (RFA) or standard surgery
* Composite outcome `dodp`: death or cancer progression within 3 years
* Confounders: `hospital`, `primary` cancer site, metastasis `position`, metastasis diameter (`diacat`).

-------------------------------------------------------------------------------

# Packages

```{r}
install.packages("MatchIt")
```
```{r}
library("haven")
library("summarytools")
library("epiDisplay")
library("magrittr")
library("MatchIt")
library("tidyverse")
options(scipen = 10, digits = 2)
```

-------------------------------------------------------------------------------

# Data exploration

We will first analyse the dataset with the usual techniques, *without* accounting for the propensity score.

1. Import the `RFA_pscore.dta` dataset and explore its variables. 
```{r}
# Load data and apply labels
rfa_pscore <- read_stata("RFA_pscore.dta") %>% mutate_if(is.labelled,as_factor)

# Explore variables
glimpse(rfa_pscore)

# Data cleaning
rfa_pscore$id <- as.factor(rfa_pscore$id)
rfa_pscore$hospital <- as.factor(rfa_pscore$hospital)
rfa_pscore$mets <- as.integer(rfa_pscore$mets)
rfa_pscore %<>% mutate(rfa = recode(
  rfa, "standard surgery" = "surgery",
  "radiofrequency ablation" = "RF ablation"
))
rfa_pscore %<>% mutate(primary = recode(primary, "small bowel/colorectal" = "bowel"))

# Data exploration
summary(rfa_pscore)
```

2. Create frequency tables of the treatment and the outcome.
```{r}
rfa_pscore %$% freq(rfa, cumul = F)
rfa_pscore %$% freq(dodp, cumul = F)
```
55% received RFA. The outcome is not rare (26%)


3.	Use logistic regression to investigate the unadjusted association between the outcome and the treatment.
Which treatment _seems to be_ associated with lower odds of death or cancer progression?
```{r}
glm(dodp ~ rfa,
    family = "binomial",
    data = rfa_pscore) %>%
  logistic.display()
```
*Solution*
Odds of outcome in RFA is half than in surgery (OR 0.55, 0.47-0.66).

However, this might be due to confounding by indication. Let's use propensity scores to explore this.

-------------------------------------------------------------------------------

At this stage you have two possible approaches:
- *Stratification*: you build a propensity score regression model, you stratify by propensity score quintiles, and you do your final analysis adjusting for propensity score quintile. This is commonly done as it is relatively simple; it's explored in points 4-11. 
- *Matching*: you fit a model that creates matched subset of your whole dataset; you perform your final analysis on this. This is more complex but more accurate; it's presented in point 12.

-------------------------------------------------------------------------------

# Propensity score estimation

4. Build a propensity score model: fit a logistic regression model with the treatment as the dependant variable and potential confounders as covariates; do _not_ include the study outcome.
Which factors are associated with undergoing RFA or surgery?


What does this tell us about our unadjusted association between RFA and outcome calculated above?

```{r}
ps_model <- glm(rfa ~ hospital + primary + position + diacat,
                family = "binomial",
                data = rfa_pscore)

logistic.display(ps_model)
# Hospital 3, moderate/difficult position, diameter < 1.5 are all associated with higher odds of undergoing RFA.
```


5. Use this model to estimate the propensity score. You use the function `predict()` that takes as an argument a regression model (so, the name of the model if you've assigned one, or you can pipe your model into this function). This function uses the model to estimate a propensity score for each row of your dataset – so you need to save these in a new variable.
```{r}
rfa_pscore$ps <- predict(ps_model, type = "response")
summary(rfa_pscore$ps)
```



6. Summarise the propensity scores by treatment group: calculate the minimum, median and maximum PS.
Which group tends to have higher propensity scores? Is this to be expected? 
```{r}
rfa_pscore %>%
  group_by(rfa) %>%
  summarise(
    "min PS" = min(ps),
    "median PS" = median(ps),
    "max PS" = max(ps)
  )
```

In which scenario would the two groups have similar propensity scores?


7. Now plot the distribution of the propensity scores by treatment group, with two histograms.
 What shape is the distribution?
```{r}
# Twin histograms with density plots
rfa_pscore %>%
  ggplot() +
  geom_histogram(aes(x = ps, y = ..density.., fill = rfa), bins = 20) +
  geom_density(aes(x = ps, y = ..density..)) +
  facet_grid(. ~ rfa) +
  theme(legend.position = "none") +
  labs(title = "Propensity Score distribution according to treatment",
       x = "Propensity Score",
       y = "Distribution density")

# Alternative 1: pooled histogram
rfa_pscore %>% ggplot(aes(ps, fill = rfa)) + geom_histogram(bins = 20) +
  labs(title = "Propensity Score distribution according to treatment",
       x = "Propensity Score",
       y = "Frequency",
       fill = "Treatment")

# Alternative 2: Box plots
rfa_pscore %>% ggplot(aes(y = rfa, x = ps)) +
  geom_boxplot(aes(fill = rfa)) +
  theme(legend.position = "none") +
  labs(title = "Propensity Score distribution according to treatment",
       x = "Propensity Score",
       y = NULL)

# Alternative 3: violin plots
rfa_pscore %>% ggplot(aes(y = rfa, x = ps)) +
  geom_violin(aes(fill = rfa)) +
  theme(legend.position = "none") +
  labs(title = "Propensity Score distribution according to treatment",
       x = "Propensity Score",
       y = NULL)
```
These graphs reveal four peaks in the distribution; which patient characteristic(s) do you think is causing these peaks?

-------------------------------------------------------------------------------

# Propensity score stratification

8. Stratify the propensity scores in quintiles.
```{r}
# Create quintiles
rfa_pscore$ps_quint <- ntile(rfa_pscore$ps, 5) %>% as.factor()

# Make sure it worked
rfa_pscore %>%
  group_by(ps_quint) %>%
  summarise(
    "min PS" = min(ps),
    "median PS" = median(ps),
    "max PS" = max(ps)
  )

# From now on, results will be slightly different to Stata. Not sure why, possibly the prediction method was different?
```

Now count how many patient had the outcome within each PS quintile:
```{r}
print("All observations")
rfa_pscore %$% ctable(ps_quint, rfa,
                      prop = "n", totals = F, headings = F)

print("Only observations with outcome")
rfa_pscore %>% filter(dodp == "yes") %$%ctable(ps_quint, rfa,
                                               prop = "n", totals = F, headings = F)
```


9. Let's investigate the association between treatment and outcome within each quintile of propensity score.

```{r}
# Stratified logistic regression dodp ~ rfa stratified by ps_quint
# aka the worst code I've ever written - but it works
# there must be a way to iterate this?
divider <- paste(rep("-", 70), collapse = "") # create a line separator

print("PS quintile 1")
glm(dodp ~ rfa, family = "binomial", data = filter(rfa_pscore, ps_quint == 1)) %>% logistic.display()

print(c(divider, "PS quintile 2"))
glm(dodp ~ rfa, family = "binomial", data = filter(rfa_pscore, ps_quint == 2)) %>% logistic.display()

print(c(divider, "PS quintile 3"))
glm(dodp ~ rfa, family = "binomial", data = filter(rfa_pscore, ps_quint == 3)) %>% logistic.display()

print(c(divider, "PS quintile 4"))
glm(dodp ~ rfa, family = "binomial", data = filter(rfa_pscore, ps_quint == 4)) %>% logistic.display()

print(c(divider, "PS quintile 5"))
glm(dodp ~ rfa, family = "binomial", data = filter(rfa_pscore, ps_quint == 5)) %>% logistic.display()

```

Similar but not the same as Stata output:
PS  1: OR 1.34 (0.84-2.14)
    2: OR 1.69 (1.20-2.36)
    3: OR 0.76 (0.52-1.10)
		4: OR 0.68 (0.44-1.05)
		5: OR 0.50 (0.20-1.25)

What do you notice in the OR in the different strata? Why do you think that's the case?

```{r 9 - Failed for loop}
output <- vector("list", 5)
for (quintile in 1:5) {
  print(quintile)
  
  result <- rfa_pscore %>% filter(ps_quint == quintile) %>%
    glm(dodp ~ rfa,
        family = "binomial",
        data = .) %>%
    logistic.display()
  result
}
```

```{r 9 - failed group by and do(glm())}
fitted_models <-
  rfa_pscore %>% group_by(ps_quint) %>% do(glm(dodp ~ rfa,
                                               family = "binomial",
                                               data = .))
```


10. Let's now calculate a single, overall estimate by adjusting for the PS stratum in a logistic regression model.
```{r}
glm(dodp ~ rfa + ps_quint,
    family = "binomial",
    data = rfa_pscore) %>% logistic.display()
# Result: Adj OR 1.05 (0.87, 1.28); Wald's test p = 0.6.
# Interpretation: Once you take the propensity score into account, there is no evidence of association between treatment and outcome.
```

Logistic regression estimates odds ratios. If we want to estimate the *risk ratio*, we need to use a different "implied link function" (not explored in ASME so far): the log link, and then exponentiate the estimates to go from log(RR) units to RR units.
```{r}
rr_model <- glm(dodp ~ rfa + ps_quint,
                family = binomial(link = "log"),  # log link goes here
                data = rfa_pscore)

# Summary in log(RR) units
summary(rr_model)

# Extracting the estimates in RR units
exp(coef(rr_model))[-1]

# Extracting the 95% CI estimates in RR units
exp(confint(rr_model))[-1, ]

# logistic.display() works but it incorrectly labels RRs as ORs
logistic.display(rr_model)

# Results: RR 1.06 (0.92-1.22)
```

Similarly, if we want to estimate the *risk difference*, we need to use the identity link.
```{r}
rd_model <- glm(dodp ~ rfa + ps_quint,
                family = binomial(link = "identity"),  # log link goes here
                data = rfa_pscore)

# Summary
summary(rd_model)

# Extracting the estimates in RD units
coef(rd_model)[-1]

# Extracting the 95% CI estimates in RD units
confint(rd_model)[-1, ]

# Result: RD -0.0038 (-0.039, 0.029)
# Stata:  RD -0.0045 (-0.041, 0.032)
```


11. (question that doesn't need coding)

-------------------------------------------------------------------------------

# Propensity score matching

We'll use functions from package MatchIt.

You need to create a matched logistic regression model, which finds pairs of observations that have similar propensity scores, but differ in the treatment status.

You do this with the function `matchit()`: your outcome goes before the ~, your propensity score model goes after it. 

Before you do this you need to remove missing values, and convert your outcome variable to integer 0/1 or logic FALSE/TRUE.

```{r}
# Omit observations with missing data
rfa_nomiss <- rfa_pscore %>% na.omit()

# Recode outcome as 0/1 integer
# There must be a quicker way of doing this with dplyr
rfa_nomiss$dodp <- as.character(rfa_nomiss$dodp)
rfa_nomiss$dodp[rfa_nomiss$dodp == "no"] <- 0
rfa_nomiss$dodp[rfa_nomiss$dodp == "yes"] <- 1
rfa_nomiss$dodp <- as.integer(rfa_nomiss$dodp)

# Fit a matching logistic regression model
mod_matched <-
  matchit(dodp ~ rfa + hospital + primary + position + diacat,
          method = "nearest",
          data = rfa_nomiss)
mod_matched
```
Note that when you call this model, you can see how many people were not matched (which STATA doesn't do)

Then, you you use this model to match the data, with `match.data()`, and you assign it to a new dataset. This will create a new variable, "distance", which represents the propensity score. This new dataset will have lost some observations, because they were not matched.

```{r}
# Create a dataset only with the matched observations. This includes a new variable called "distance", which is the propensity score.
rfa_matched <- match.data(mod_matched)
dim(rfa_matched)
```

Now, _I guess_ we can fit our final model to this matched dataset?
Like this?
```{r}
rd_matched_model <- glm(dodp ~ rfa,
                        family = binomial(link = "identity"),
                        data = rfa_matched)

# Summary
summary(rd_matched_model)

# Extracting the estimates in RD units
coef(rd_matched_model)[-1]

# Extracting the 95% CI estimates in RD units
confint(rd_matched_model)[-1,]

# Results: RD +0.026 (-0.047, +0.052)
# Stata:   RD -0.028 (-0.089, +0.033)
```
How do you interpret these risk difference results?


13. Compare the results from stratification and the results from matching.

- Which risk difference has the wider confidence interval? Why do you think that is?
- Are the two risk difference estimates different? Would you expect them to be?

-------------------------------------------------------------------------------

## References
1. ARM Propensity Score practical
2. https://sejdemyr.github.io/r-tutorials/statistics/tutorial8.html