---
title: "Missing data: Exploring the shortcomings of ad-hoc methods and creating multiple imputations"
subtitle: "R 4 ARM"
author: Julian Matthewman [link](https://github.com/julianmatthewman)
---

-------------------------------------------------------------------------------

## Prerequisites
This is an attempt at translating the Stata practical "Exploring the shortcomings of ad-hoc methods and creating multiple imputations in Stata" from LSHTM's Advanced Research Methods module into R.

## What you will learn
1. How to implement the missing category method and complete record analysis
2. How to do multiple imputation for a binary variable
3. The problems with mean and regression mean imputation for continuous variables with missing values
4. How to do simple multiple imputation for a continuous variable

-------------------------------------------------------------------------------

## Data management

```{r setup}
#Either have the files you want to read in the same directory as this file, or change the working directory here:
knitr::opts_knit$set(root.dir = ("/Users/Julian/Dropbox/Documents/MSc Epidemiology/Advanced Research Methods/Tuesday am Missing Data-20200421/Home practical/"))
#This is a way of setting the working directory when working with chunks in R notebooks. Otherwise we would have to define the working directory separately for each chunk.
```

```{r message=FALSE, warning=FALSE}
# Load packages
library(haven)
library(magrittr)
library(epiDisplay)
library(ggplot2)
library(dplyr)

# Limit significant digits to 2, remove scientific notation
options(digits = 2, scipen = 999)
```

```{r}
# Data import
# Etiher have the data files in the same directory as this R notebook or use setwd("yourdirectory/ASMEdata2020")
full <- read_dta("prac2full.dta")
miss <- read_dta("prac2miss.dta")
```

Variable |  Missing Observations? | Details
id          n                       Unique individual identifier
sex         n                       Sex (0 = male, 1 = female)
smok0       y                       Smoking status (1 = current smoker, 0 = non-smoker)
x0          n                       A continuous variable measured at baseline
y0          n                       A continuous variable measured at baseline
y1          y                       y0 measured at a follow-up time
t           n                       Time to death or censoring in years (censoring occurs at 10y)
d           n                       Indicator of death (1 = died, 0 = censored)


```{r}
glimpse(full)
glimpse(miss)
summary(miss)
```

# 4.Estimating the mean of y1
## 4.1 Full data analysis
Using the complete data, estimate the mean of y1.
```{r}
mean(full$y1)
```

## 4.2 Check for missing values
```{r}
summary(miss$y1)
#Create a variable M_y1 equal to 0 if y1 is observed and 1 if y1 is missing:
miss <- miss %>% mutate(M_y1 = case_when(
    is.na(y1) == TRUE ~ 1,
    is.na(y1) == FALSE ~ 0
))
```

## 4.3 Mean imputation
Create a new variable called y1meanimp containing all the values from y1 and replacing all the missing values of y1 with the mean of the observed values of y1.
```{r}
miss <- miss %>% mutate(y1meanimp = case_when(
    is.na(y1) == TRUE ~ mean(miss$y1, na.rm = TRUE),
    is.na(y1) == FALSE ~ y1
))
```

Create a scatterplot and a quantile-quantile plot of the imputed values:
```{r}
ggplot(miss) +
    geom_point(aes(x=x0, y=y1meanimp))
ggplot(miss) +
    geom_qq(aes(sample = y1meanimp))
```

Calculate the mean, and corresponding confidence interval for y1meanimp
```{r}
ci(miss$y1meanimp)
```

## 4.4 Regression imputation
Now we use a regression model of y1 on y0 and x0 to impute the missing values of y1.

First, look at the results of a regression model of y1 on y0 and x0, and explain why y0 is a good candidate variable for imputing missing y1, but x0 may not be so useful.
```{r}
glm(y1 ~ y0, data = miss) %>% summary()
```

Now we use regression to impute the missing values of y1:
```{r}
model1 <- glm(y1 ~ y0 + x0, data = miss)

miss$y1regimp <- NA
miss$y1regimp[is.na(miss$y1)] <-  predict(model1, newdata = miss[is.na(miss$y1),])
#predict value from model1. newdata is the data frame in which to look for variables with which to predict. 
miss <- miss %>% mutate(y1regimp = case_when(
    is.na(y1) == FALSE ~ y1,
    is.na(y1) == TRUE ~ y1regimp
))
```

Now we use the commands from the previous subsection to inspect the imputed data. Are these values more plausible?
```{r}
ggplot(miss) +
    geom_point(aes(x=x0, y=y1regimp))
ggplot(miss) +
    geom_qq(aes(sample = y1regimp))
ci(miss$y1regimp)
glm(y1regimp ~ 1, data=miss) %>% summary() #to check if it is the same as in the Stata solutions
```

## 4.5 Multiple imputation
Now we will try multiple imputation.
As discussed above, it is useful to include variables that are either:
1. both predictive of missingness and good predictors of missing y1 values or 
2. good predictors of missing y1 values.
Variables that only predict missingness do not need to be included in the imputation model.


```{r}
library(mice)
#If we want to change the "settings" of mice, we can use the following commented code:
#init = mice(miss) 
#meth = init$method
#predM = init$predictorMatrix
#remove variables as predictors
#predM[, c("id", "sex", "smok0", "t", "d", "M_y1", "y1meanimp", "y1regimp")]=0
#skip a variable from imputation
#meth[c("smok0")]=""
#specifying imputation methods
#meth[c("y1")]="pmm" 

#run mulitple imputation
#m = number of imputations to be done
#seed = seed for random number generator, i.e.: with this seed should get the same results
#method = select imputation method for each variable (here is left blank for the first two variables)
#"pmm" or predicitve mean matching is the default method
imp <- mice(miss[3:5], m=10, seed = 9861, method = c("", "", "pmm")) 
#When using "norm.predict" as the imputation method it gives us the exact same results as above, when we did the regression mean imputation. I would assume that "norm.predict" should be equivalent to mi impute regress in stata; no idea why the results are different.

miss$y1mi <-  complete(imp)$y1

ggplot(miss) +
    geom_point(aes(x=x0, y=y1mi))
ggplot(miss) +
    geom_qq(aes(sample = y1mi))

ci(miss$y1mi)
glm(y1mi ~ 1, data=miss) %>% summary()
```


# 5.Cox model
We will skip this section; I can't get the code to work  ¯\_(ツ)_/¯ 

#6. Your results
```{r}
results <- bind_rows(ci(full$y1),
                     ci(miss$y1),
                     ci(miss$y1meanimp),
                     ci(miss$y1regimp),
                     ci(miss$y1mi)) %>% 
    bind_cols(tibble(method = c("Complete Data", "Complete Cases", "Mean Imp.", "Reg. Imp.", "Multiple Imp.")))
results
    
ggplot(results) +
    geom_point(aes(x=method, y=mean)) +
    geom_errorbar(aes(x=method, y=mean, ymin = lower95ci, ymax = upper95ci)) +
    coord_flip()
```

We can see that:
* Complete records and mean imputation led to very biased results.
* Regression imputation and even more Mean imputation have too short intervals; shorter than the actual data
* See practical excercises for the rest


```{r}
#Comparing qq plots, don't know if this makes sense
ggplot(full, aes(sample = y1)) +
    geom_qq() +
    geom_qq_line() +
    scale_y_continuous(limits = c(-40, 40))
ggplot(miss, aes(sample = y1mi)) +
    geom_qq() +
    geom_qq_line() +
    scale_y_continuous(limits = c(-40, 40))
ggplot(miss, aes(sample = y1regimp)) +
    geom_qq() +
    geom_qq_line() +
    scale_y_continuous(limits = c(-40, 40))
ggplot(miss, aes(sample = y1meanimp)) +
    geom_qq() +
    geom_qq_line() +
    scale_y_continuous(limits = c(-40, 40))
```



















#############################
# Skipped section (not working due to porblem with nelsonaalen function)
#############################

## 5. Cox model with an incomplete binary covariate
### 5.1 Full data analysis
We begin by analysing the full dataset; we fit a Cox model relating survival to the covariates sex, smok0, y0 and x0.
```{r}
library(survival)
cox <- coxph(Surv(full$t, full$d) ~ sex + smok0 + y0 + x0, data = full)
#Going into survival analysis is a whole other (complicated) topic; suffice it to say that to fit a Cox model we need a survival object which we can create with the Surv() function from the survival package. If you have a look at Surv(full$t, full$d) you can see that it's basically a list of survival time, with "+" meaning a censored observation.
summary(cox) #same results as in Stata
```

### 5.2 Complete record analysis
```{r}
library(survival)
cox <- coxph(Surv(miss$t, miss$d) ~ sex + smok0 + y0 + x0, data = miss)
summary(cox) #same results as in Stata
```


### 5.3 Missing category method
As a way to make use of the available data in other covariates, create a ‘missing’ category for smoking. We then fit the Cox model again with the new variable.
```{r}
miss <- miss %>% mutate(smok0mc = case_when(is.na(smok0) == TRUE ~ 2,
                                            is.na(smok0) == FALSE ~ smok0))

cox <- coxph(Surv(miss$t, miss$d) ~ sex + as.factor(smok0mc) + y0 + x0, data = miss)
summary(cox) #same results as in Stata
```
For x0, the missing category method does not seem unreasonable, but it is rather biased for smok0.

### 5.4 Multiple imputation
Note that any association between the incomplete covariate and outcome needs to be represented in the imputed data, and so it is important to include the outcome as a covariate of the imputation model. With survival data this is most conveniently done by including the Nelson- Aalen estimate of the cumulative hazard and the censoring indicator in the imputation model.
```{r}
nelsonaalen(miss, t, d) #there is an error with this function "Time variable is not numeric"
```

